---
title: "Crude mortality"
author: "Göran Broström"
date: "18 June 2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "", fig.height = 4, fig.path = "figs/")
```

# Introduction

Just basic variables, birthdate, deathdate (if any), sex, snuffing, time for 
VIP.

# Read data and keep essentials

```{r readdata}
vip <- readRDS("Rdata/vip.rds")
cr <- vip[, c("LopNr", "provdat", "fyear", "sex", "DODSDAT", "h1a", "h4")]
cr$DODSDAT <- as.Date(as.character (cr$DODSDAT), format = "%Y%m%d")
cr$sex <- factor(cr$sex, labels = c("male", "female"))
summary(cr)
```

# Select never-smokers

```{r neversmokers}
#cr <- cr[!is.na(cr$h1a), ]
#cr$h1a <- NULL
```

# Calculate ages

```{r ages}
cr$enter <- as.numeric(cr$provdat - cr$fyear) / 365.25
cr$event <- !is.na(cr$DODSDAT)
lastSeen <- cr$DODSDAT
lastSeen[is.na(lastSeen)] <- as.Date("2015-12-31")
cr$exit <- as.numeric(lastSeen - cr$fyear) / 365.25
summary(cr)
```

OK, but need to remove individuals without information on snus habits, `r sum(is.na(cr$h4))`
persons.

```{r rmna}
#cr <- cr[!is.na(cr$h4), ]
#with(cr, table(h4))
```



```{r agprov, fig.cap = "Age distribution at baseline."}
with(cr, hist(enter, xlab = "Age", main = "", breaks = 160))
```

The distribution of *age at prov*, see Figure \@ref(fig:agprov).

# Coding snus use

What is *h4 == 9*? Probably a missing-value-code, so we remove them.

```{r h49}
cr <- cr[cr$h4 != 9, ]
```

And we create a new variable `snuffer`, which is a recoding of `h4`:

```{r recode}
cr$snuffer <- cut(cr$h4, c(0, 1.5, 2.5, 6.5), labels = c("never", "previous", "current"))
x <- with(cr, table(snuffer, event))
round(with(cr, prop.table(x, margin = 1)) * 100, 2)
```

Somewhat surprising that there are relatively more deaths among never-snuffers,
but it may be explained by the fact that they are older at `provdat`:

```{r agebysnuff}
with(cr, tapply(enter, snuffer, mean))
```
As expected.

# Introducing age cohorts

Ages at `provdat` are heaping around ages 30, 40, 50, and 60. We create a `cohort`
accordingly.

```{r cohort}
cr$cohort <- cut(cr$enter, c(0, 35, 45, 55, 65), 
                 labels = c("age_30", "age_40", "age_50", "age_60"))
with(cr, table(cohort, event, sex))
```


Lets look at some cumulative hazards, with `age` as baseline time scale.

```{r age, fig.cap = "Snuffers, age as time scale."}
library(eha)
fit0 <- coxreg(Surv(enter, exit, event) ~ snuffer, data = cr)
##plot(fit0, xlab = "Age")
summary(fit0)
```

Also as expected. With *time since provdat* as baseline time scale we get

```{r sincenon, fig.cap = "Snuffers, duration as time scale, not 'adjusted' for age."}
fit1 <- coxreg(Surv(exit - enter, event) ~ strata(snuffer), data = cr)
plot(fit1, xlab = "Duration (years)")
```



```{r since, fig.cap = "Snuffers, duration as time scale, 'adjusted' for age (ref 40)."}
fit2 <- coxreg(Surv(exit - enter, event) ~ I(enter - 40) + strata(snuffer), data = cr)
plot(fit2, xlab = "Duration (years)")
```

Also as expected, since we have not adjusted for age. Let's do that, see Figure \@ref(fig:since).


Adjusting also for `sex`:

```{r sincesex, fig.cap = "Snuffers, duration as time scale, 'adjusted' for age (ref 40) and sex."}
par(lwd = 1.5)
fit3 <- pchreg(Surv(exit - enter, event) ~  strata(snuffer), 
               cuts = seq(0, 25, by = 5), 
               data = cr[cr$sex == "male" & cr$cohort == "age_60", ])
plot(fit3, , fn = "haz", xlab = "Duration (years)")
```
