---
title: "Snus and mortality"
author: "Göran Broström"
date: "`r Sys.time()`"
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 3
  bookdown::word_document2:
    toc: no
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
citation_package: natbib
classoption: titlepage
bibliography: bio.bib
titlepage: yes
biblio-style: apalike
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{a4wide}
#- \usepackage[tablesonly]{endfloat}
#- \renewcommand{\efloatseparator}{\mbox{}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
source("R/ltxCoef3.R")
library(eha)
```

# Introduction

We want to study the effect of snus use on overall mortality, controlling for age, sex,
"frukt och grönt", education, and income.

# Data

All the data we need are stored in the file *vip.rds*, which is stored in the subdirectory *Rdata*.
We read it and at the same time select the variables we need. First, construct a vector that 
contains the names of the variables:

```{r varlist}
varnames <- c("LopNr", "provdat", "birthdate", "enter", "exit", "event", 
              "snususe", "sex", "green", "education", "uincome")
```

Then we read *vip.rds* and make the selection at the same time. We give the selected subset the name *snuff* 
(or whatever, not important).

```{r readdata}
snuff <- readRDS("Rdata/vip.rds")[, varnames]
```

Note the way selection is made: The "[, ]" is the key, rows are specified to the left of the comma, 
and columns to the right. Specifying no rows (leaving empty as here) selects all rows (all persons).

Let us look at our subsample, two help functions, *str* and *summary*, are useful:

```{r str}
str(snuff)
```

*str* is shorthand for *structure*, and that is what the function shows: A list of all variables, their *type* (num, Date, logic, Factor),
which is helpful to know. The first few values of each variable is also presented.

Then the *summary* function*:
```{r summary}
summary(snuff)
```

The variable *enter* is age at *provdat* (essentially *provdat* minus *birthdate*, expressed in years), 
*exit* is age at "lost to follow-up* and *event* is a logical, *TRUE* if lost to follow-up is due to death,
*FALSE* otherwise (could also be coded as 1 and 0, respectively).

A thing to worry about is missing values (*NA*). Here I simply choose to remove all cases with any missing value:
The function *na.omit* does that trick:

```{r omit}
snuff1 <- na.omit(snuff)
NROW(snuff1)
summary(snuff1)
```

Around 8000 individuals are lost due to this, leaving 43847 ones. More seriously, we lost a great number of deaths!
Maybe we should make *NA* a separate category for *green* and *education* (luckily we kept the original *snuff*):

```{r catna}
snuff$green <- as.character(snuff$green) # Probably necessary ...
snuff$green[is.na(snuff$green)] <- "unknown"
snuff$green <- factor(snuff$green)
## And education:
snuff$education <- as.character(snuff$education)
snuff$education[is.na(snuff$education)] <- "Unknown"
snuff$education <- factor(snuff$education)
## Remove missing:
snuff1 <- na.omit(snuff)
## Check result:
summary(snuff1)
```

A little bit "complicated", but a rather frequent request, so it may be worthwhile to 
write a general function \ldots

Anyway, we saved the deaths!

# Cox regression

The first important question is to choose *start event* (time scale). Two possible choices: (i) Birth, and (ii) Prov.
Since *age* is such a strong determinant of mortaltity, often (i) is chosen, but chosing (ii) also has merit 
(keeping age at provat as a covariate).

## Birth as start event

See Table \@ref(tab:startbirth).

```{r startbirth, results = "asis"}
fit <- coxreg(Surv(enter, exit, event) ~ snususe + sex + green + education + I(uincome - 1550), 
              data = snuff1)
dr <- drop1(fit, test = "Chisq")
ltxCoef3(fit, dr = dr, digits = 4, caption = "Birth as start event.", label = "tab:startbirth")
```

A few things to note:

1.   The only effect of *green* seems to be that the *unknown* group has a very low mortality, relatively speaking.

2.   For *education*, the effect of *unknown* is reversed, an extremely high mortality.

3.   For *income* the CI is (1.0000, 1.0000), but it is significantly different from 1! This is a scale problem,
easy to solve.

4.   Center continuous covariates (income).

## Prov as start event

See Table \@ref(tab:startbirth).

```{r startprov, results = "asis"}
fit <- coxreg(Surv(exit - enter, event) ~ I(enter - 45) + snususe + sex + green + 
                 education + I((uincome - 1550)/ 100), 
              data = snuff1)
dr <- drop1(fit, test = "Chisq")
ltxCoef3(fit, dr = dr, digits = 4, caption = "Prov as start event.", label = "tab:startprov")
```

Note that *enter* in Table \@ref(tab:startprov) is *age at provtagning*. Otherwise similar 
remarks as in the previous case. *Except* that we note that income now is scaled properly,
and therefore the CI is useful \ldots

### A graph of snuff mortality

```{r plott, fig.cap = "Cumulative hazards after prov by snus use."}
fit <- coxreg(Surv(exit - enter, event) ~ I(enter - 45) + strata(snususe) + sex + 
                 green + education + I(uincome - 1550), 
              data = snuff1)
plot(fit, main = "", xlab = "Time after prov", col = c("black", "blue", "red"))
```

Let's end the follow-up at age 22:

```{r resplott, fig.cap = "Cumulative hazards after prov by snus use, follow-up till 22."}
snuff1$exit2 <- snuff1$exit - snuff1$enter
snuff1$enter2 <- 0
snuff2 <- age.window(snuff1, c(0, 22), surv = c("enter2", "exit2", "event"))
fit <- coxreg(Surv(exit - enter, event) ~ I(enter - 45) + strata(snususe) + sex + 
                 green + education + I(uincome - 1550), 
              data = snuff2)
plot(fit, main = "", xlab = "Time after prov", col = c("black", "blue", "red"))
```

# Conclusion

This was an intro to how an analysis may start up. Saw already some questions that need answers. 
