---
title: "Gompertz"
author: "Göran Broström"
date: "`Sys.Date()`"
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: no
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::word_document2:
    toc: no
    toc_depth: 2
citation_package: natbib
##classoption: titlepage
bibliography: snus2.bib
titlepage: no
biblio-style: apalike
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{sectsty}
- \subsectionfont{\normalfont\itshape}
---

# Introduction

I'm interested in trying to fit the *Gompertz* distribution to the mortality analyses 
with the *vip* data. Reason is that it is well-known that the Gompertz distribution
fits adult mortality very well in general, and we can draw stronger conclusions
with a parametric model compared to a semi-parametric one like Cox regression.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data

```{r readvip}
vip <- readRDS("Rdata/vip.rds")
library(eha)
```

# Analysis of a simple model

## Age 30 at prov

```{r coxreg30, fig.cap = "Gompertz and non-parametric cumulative hazards, age 30."}
vip30 <- vip[vip$alder.VIP == 30, ]
fit.c <- coxreg(Surv(exit, event) ~ 1, data = vip30)
fit.g30 <- phreg(Surv(exit, event) ~ 1, data = vip30, dist = "gompertz", param = "rate")
check.dist(fit.c, fit.g30, main = "")
```



```{r coxreg40, fig.cap = "Gompertz and non-parametric cumulative hazards, age 40."}
vip40 <- vip[vip$alder.VIP == 40, ]
fit.c <- coxreg(Surv(exit, event) ~ 1, data = vip40)
fit.g40 <- phreg(Surv(exit, event) ~ 1, data = vip40, dist = "gompertz", param = "rate")
check.dist(fit.c, fit.g40, main = "")
```



```{r coxreg50, fig.cap = "Gompertz and non-parametric cumulative hazards, age 50."}
vip50 <- vip[vip$alder.VIP == 50, ]
fit.c <- coxreg(Surv(exit, event) ~ 1, data = vip50)
fit.g50 <- phreg(Surv(exit, event) ~ 1, data = vip50, dist = "gompertz", param = "rate")
check.dist(fit.c, fit.g50, main = "")
```



```{r coxreg60, fig.cap = "Gompertz and non-parametric cumulative hazards, age 60."}
vip60 <- vip[vip$alder.VIP == 60, ]
fit.c <- coxreg(Surv(exit, event) ~ 1, data = vip60)
fit.g60 <- phreg(Surv(exit, event) ~ 1, data = vip60, dist = "gompertz", param = "rate")
check.dist(fit.c, fit.g60, main = "")
```


## Check Gomperts baseline parameters

```{r checkpar}
out <- matrix(0, ncol = 2, nrow = 4)
out[1, ] <- coef(fit.g30)
out[2, ] <- coef(fit.g40)
out[3, ] <- coef(fit.g50)
out[4, ] <- coef(fit.g60)
rownames(out) <- paste0("age.", 10 * (3:6))
colnames(out) <- c("Rate", "Level")
out[, 2] <- exp(out[, 2])
source("R/tbl.R")
out <- round(out, 4)
tbl(out, rownames = TRUE, caption = "Gompertz estimated parameters")
```

## Comparing models

We want to compare a model stratified by education with a model with education included 
as a factor in the Gompertz setting.

```{r factorvstrata}
fit.f <- phreg(Surv(exit, event) ~ age + education, data = vip40, dist = "gompertz")
fit.s <- phreg(Surv(exit, event) ~ age + strata(education), data = vip40, dist = "gompertz")
anova(fit.f, fit.s)
```

