---
title: "Gompertz"
author: "Göran Broström"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: no
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::word_document2:
    toc: no
    toc_depth: 2
citation_package: natbib
##classoption: titlepage
bibliography: snus2.bib
titlepage: no
biblio-style: apalike
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{sectsty}
- \subsectionfont{\normalfont\itshape}
- \usepackage[figuresonly]{endfloat}
- \renewcommand{\efloatseparator}{\mbox{}}
---

# Introduction

I'm interested in trying to fit the *Gompertz* distribution to the mortality analyses 
with the *vip* data. Reason is that it is well-known that the Gompertz distribution
fits adult mortality very well in general, and we can draw stronger conclusions
with a parametric model compared to a semi-parametric one like Cox regression.

The Gompertz distribution is characterized by the *hazard function*

\begin{equation*}
h(t; \alpha, \beta) = \alpha e^{\beta t}, \quad \alpha, t > 0, -\infty < \beta < \infty,
\end{equation*}

where $\alpha$ is the *level* and $\beta$ the *growth rate*. If $\beta$ is *constant*, 
this constitutes a family of *proportional hazards* distributions as $\alpha$ varies.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.height = 4)
```

# Data

Reading in *vip.rda*.

```{r readvip}
vip <- readRDS("Rdata/vip.rds")
library(eha)
par(lwd = 2)
```

# Analysis of a simple model

We fit a simple Gompertz distribution to each of the four age groups (no covariates) 
and judge the fits graphically, see Figures 1--4. As you can see, there are excellent 
fits in all cases, suggesting that we should do all the analyses with the Gompertz 
rather than Cox model (both are proportional hazards models, but the Gompertz can also
be twisted into an *accelerated failure time model*).

We want to see if the four estimated growth rates may be considered equal, 
implying *proportional hazards*. 


```{r coxreg30, fig.cap = "Gompertz and non-parametric cumulative hazards, age 30."}
vip30 <- vip[vip$alder.VIP == 30, ]
fit.c <- coxreg(Surv(exit, event) ~ 1, data = vip30)
fit.g30 <- phreg(Surv(exit, event) ~ 1, data = vip30, 
                 dist = "gompertz", param = "rate")
check.dist(fit.c, fit.g30, main = "")
```



```{r coxreg40, fig.cap = "Gompertz and non-parametric cumulative hazards, age 40."}
vip40 <- vip[vip$alder.VIP == 40, ]
fit.c <- coxreg(Surv(exit, event) ~ 1, data = vip40)
fit.g40 <- phreg(Surv(exit, event) ~ 1, data = vip40, 
                 dist = "gompertz", param = "rate")
check.dist(fit.c, fit.g40, main = "")
```



```{r coxreg50, fig.cap = "Gompertz and non-parametric cumulative hazards, age 50."}
vip50 <- vip[vip$alder.VIP == 50, ]
fit.c <- coxreg(Surv(exit, event) ~ 1, data = vip50)
fit.g50 <- phreg(Surv(exit, event) ~ 1, data = vip50, 
                 dist = "gompertz", param = "rate")
check.dist(fit.c, fit.g50, main = "")
```



```{r coxreg60, fig.cap = "Gompertz and non-parametric cumulative hazards, age 60."}
vip60 <- vip[vip$alder.VIP == 60, ]
fit.c <- coxreg(Surv(exit, event) ~ 1, data = vip60)
fit.g60 <- phreg(Surv(exit, event) ~ 1, data = vip60, 
                 dist = "gompertz", param = "rate")
check.dist(fit.c, fit.g60, main = "")
```


## Check Gomperts baseline parameters

```{r checkpar}
out <- matrix(0, ncol = 2, nrow = 4)
out[1, ] <- coef(fit.g30)
out[2, ] <- coef(fit.g40)
out[3, ] <- coef(fit.g50)
out[4, ] <- coef(fit.g60)
rownames(out) <- paste0("age.", 10 * (3:6))
colnames(out) <- c("Rate", "Level")
out[, 2] <- exp(out[, 2])
source("R/tbl.R")
out <- round(out, 4)
tbl(out, rownames = TRUE, caption = "Gompertz estimated parameters.")
```

Table \@ref(tab:checkpar) shows that the rate is increasing with age (exception: at 50),
and a formal test of equal rates is firmly rejected ($p$-value 0.000015). And this means that the hypothesis 
of proportional hazards is rejected, see the next section. 

## Comparing models

We want to compare a model stratified by *alder.VIP* with a model with alder.VIP included 
as a factor in the Gompertz setting.

```{r factorvstrata, fig.cap = "Hazard functions for age groups."}
fit.f <- phreg(Surv(exit, event) ~ alder.VIP, data = vip, 
               dist = "gompertz", param = "rate")
fit.s <- phreg(Surv(exit, event) ~ strata(alder.VIP), data = vip, 
               dist = "gompertz", param = "rate")
##anova(fit.f, fit.s)
ttt <- 2 * (fit.s$loglik[2] - fit.f$loglik[2])
pval <- pchisq(ttt, df = 3, lower.tail = FALSE)
par(lwd = 2, las = 1)
plot(fit.s, fn = "haz", printLegend = "topleft", col = 1:4, main = "",
     xlab = "Time after test (years)")
par(lwd = 1)
abline(h = 0)
```

Figure \@ref(fig:factorvstrata) illustrates the results in Table \@ref(tab:checkpar).
