---
title: "Interactions"
author: "Göran Broström"
date: "4/7/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(eha)
source("R/ltx2.R")
vip <- readRDS("Rdata/vip.rds")
```

# livingalone and inc.90

```{r withsnus}
fit <- coxreg(Surv(exit, event) ~ livingalone * snususe + strata(alder.VIP), 
              data = vip)
drop1(fit, test = "Chisq")
```

Nothing remarkable.

```{r withinc}
fit <- coxreg(Surv(exit, event) ~ livingalone * inc.90 + snususe + strata(alder.VIP), 
              data = vip)
drop1(fit, test = "Chisq")
```

Oh dear, a very strong interaction, let's do separate analyses for livingalone categories:


```{r livalone, results = "asis"}
fit <- coxreg(Surv(exit, event) ~ snususe + inc.90 + strata(alder.VIP), 
              data = vip[vip$livingalone == "Yes" & !is.na(vip$inc.90), ])
ltx2(fit, caption = "Living alone.")
```


```{r notlivalone, results = "asis"}
fit <- coxreg(Surv(exit, event) ~ snususe + inc.90 + strata(alder.VIP), 
              data = vip[vip$livingalone == "No" & !is.na(vip$inc.90), ])
ltx2(fit, caption = "Not living alone.")
```

Much stronger (negative) effects of both snuffing and low income for those 
who live alone.

Isn't this a finding worth mentioning?

# livingalone and sex

```{r sexsnus}
fit <- coxreg(Surv(exit, event) ~ snususe + livingalone * sex + strata(alder.VIP), 
              data = vip)
drop1(fit, test = "Chisq")
```

Right, a significant interaction effect. Reanalyse by sex:

```{r womsnus, results = 'asis'}
fit <- coxreg(Surv(exit, event) ~ snususe + livingalone + strata(alder.VIP), 
              data = vip[vip$sex == "Women", ])
ltx2(fit, caption = "Women.")
```

```{r mensnus, results = 'asis'}
fit <- coxreg(Surv(exit, event) ~ snususe + livingalone + strata(alder.VIP), 
              data = vip[vip$sex == "Men", ])
ltx2(fit, caption = "Men.")
```

# livingalone and cholesterol

