---
title: "Proportional hazards, men"
author: "Göran Broström"
date: "4/5/2022"
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: no
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::word_document2:
    toc: no
    toc_depth: 2
citation_package: natbib
##classoption: titlepage
bibliography: snus2.bib
titlepage: no
biblio-style: apalike
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{sectsty}
- \subsectionfont{\normalfont\itshape}
---

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
vip <- readRDS("Rdata/vip.rds")
vip <- vip[vip$sex == "Men", ]
##
library(eha)
library(survival)
source("R/tbl.R")
source("R/ltx2.R")
```

# Check of proportional hazards assumption

Essentially two ways of checking proportional hazards with real data.

## Graphical check

Works well with a categorical (*factor*) covariate. We take *alder.VIP*
as an example, see Figure \@ref(fig:gralder).

```{r gralder, fig.cap = "Stratification wrt alder.VIP, men. log-log scale."}
fit <- coxreg(Surv(exit, event) ~ strata(alder.VIP), data = vip)
plot(fit, col = 1:4, lty = c(1, 2, 4, 5),
     printLegend = "topleft", fn = "loglog")
```

On the *log-log* scale, the curves should have a constant vertical distance to each other,
bu that seems less plausible, for instance are the curves corresponding to ages 30 and 40
crossing.


## Formal test

We use the function *cox.zph* in the *survival* package. And we have to replace the call to
*coxreg* with a call to *coxph*. They do the same job, but *coxreg* cannot cope with *cox.zph*. 
So, see Table \@ref(tab:zph). It shows the results of tests of null hypotheses of proportionality,
and we can see that the hypothesis concerning *alder.VIP* is clearly rejected (very small *p*-value).
On the other hand, no big problem regarding *sex*, and the *GLOBAL* conclusion is that the model 
as a whole does not fulfil the requirement of proportionality.

```{r zph, results = 'asis'}
fit <- coxph(Surv(exit, event) ~ alder.VIP, data = vip)
zz <- cox.zph(fit)
tbl(zz$table, caption = "Test of proportionality, men.", 
    rownames = TRUE)
```

# Choice of age dependence

We have no less than *three* variables referring to age in some way or the other, *exit*, 
*alder_VIP* and *late*. We arrange them so that *exit* is the fundamental duration variable
(measuring time between test and death), *alder.VIP* is age test group(time fixed), and 
*late* measures the exact time between alder.VIP and age at test.

Like this, see Table \@ref(tab:basic).

```{r basic, results = 'asis'}
fit.basic <- coxreg(Surv(exit, event) ~ snususe  + us.ar + late 
                    + strata(alder.VIP), data = vip)
fit.ph <- coxph(Surv(exit, event) ~ snususe + us.ar + late 
                    + strata(alder.VIP), data = vip)
ltx2(fit.basic, label = "tab:basic", caption = "Basic model, men.")
```

# Put on more covariates

Like this, see Table \@ref(tab:basic2).

```{r basic2, results = 'asis'}
vip0 <- vip[!is.na(vip$education) &
               !is.na(vip$inc.90), ]
fit.basic <- coxreg(Surv(exit, event) ~ snususe  +  livingalone +
                       education + inc.90 +
                       us.ar + 
                       Boende + late 
                    + strata(alder.VIP), data = vip0)
fit.ph <- coxph(Surv(exit, event) ~ snususe + livingalone +
                   education + inc.90 +
                   us.ar + Boende + 
                   late 
               + strata(alder.VIP)
                , data = vip0[vip0$alder.VIP == "40", ])
ltx2(fit.basic, label = "tab:basic2", caption = "Stratified by alder.VIP, men.")
```


A test of proportionality, see Table \@ref(tab:zph2).

```{r zph2, results = 'asis'}
zz <- cox.zph(fit.ph)
tbl(zz$table, caption = "Test of proportionality.", 
    rownames = TRUE)
```

Seems OK, with a small reservation for *inc.90*. The *GLOBAL* conclusion is OK.

Not so much significance to talk about for *snususe*, when more covariates are added.
However, we start by looking at each stratum  (age group) separately.

## Age group 30

```{r basic230, results = 'asis'}
fit.basic <- coxreg(Surv(exit, event) ~ snususe  + livingalone +
                       education + inc.90 +
                       us.ar + 
                       Boende + late 
                    , data = vip0[vip0$alder.VIP == "30", ])
fit.ph <- coxph(Surv(exit, event) ~ snususe + livingalone +
                   education + inc.90 +
                   us.ar + Boende + 
                   late 
               
                , data = vip0[vip0$alder.VIP == "30", ])
ltx2(fit.basic, label = "tab:basic230", caption = "alder.VIP = 30, men.")
```

## Age group 40

```{r basic240, results = 'asis'}
fit.basic <- coxreg(Surv(exit, event) ~ snususe + inc.90 +
                       us.ar + livingalone +
                       Boende + late, 
                    data = vip0[vip0$alder.VIP == "40", ])
fit.ph <- coxph(Surv(exit, event) ~ snususe + livingalone +
                   education + inc.90 +
                   us.ar + Boende + 
                   late 
               
                , data = vip0[vip0$alder.VIP == "40", ])
ltx2(fit.basic, label = "tab:basic240", caption = "alder.VIP = 40, men.")
```

## Age group 50

```{r basic250, results = 'asis'}
fit.basic <- coxreg(Surv(exit, event) ~ snususe  + livingalone +
                       education + inc.90 +
                       us.ar + 
                       Boende + late 
                    , data = vip0[vip0$alder.VIP == "50", ])
fit.ph <- coxph(Surv(exit, event) ~ snususe  + livingalone +
                   education + inc.90 +
                   us.ar + Boende + 
                   late 
         
                , data = vip0[vip0$alder.VIP == "50", ])
ltx2(fit.basic, label = "tab:basic250", caption = "alder.VIP = 50, men.")
```

## Age group 60

```{r basic260, results = 'asis'}
fit.basic <- coxreg(Surv(exit, event) ~ snususe  + livingalone +
                       education + inc.90 +
                       us.ar + 
                       Boende + late 
                    , data = vip0[vip0$alder.VIP == "60", ])
fit.ph <- coxph(Surv(exit, event) ~ snususe + livingalone +
                   education + inc.90 +
                   us.ar + Boende + 
                   late 
   
                , data = vip0[vip0$alder.VIP == "60", ])
ltx2(fit.basic, label = "tab:basic260", caption = "alder.VIP = 60, men.")
```

