---
title: "Tables and graphs"
author: "Göran Broström"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: no
    toc_depth: 3
  bookdown::word_document2:
    toc: no
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
citation_package: natbib
##classoption: titlepage
bibliography: bio.bib
##titlepage: no
biblio-style: apalike
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{a4wide}
- \usepackage[figuresonly]{endfloat}
- \renewcommand{\efloatseparator}{\mbox{}}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "")
```

# Tables

## stargazer

```{r starg, results='asis', message = FALSE}
vip <- readRDS("Rdata/vip.rds")
names(vip) <- gsub("_", ".", names(vip))
## Warning: The following is not necessary with latest version of vip!
vip$exit <- vip$exit - vip$enter
vip$age <- vip$enter
vip$enter <- NULL
kont <- vip[, c("income.90", "age", "exit", "bmi", "sex", "alder.VIP", "snususe")]
library(stargazer)
stargazer(kont, title = "vip data.", header = FALSE, summary = TRUE, 
          label = "tab:starg")
```

The columns in Table \@ref(tab:starg) are the default ones chosen by *stargazer*. Can
be manipulated.


<!--
This is out-commented code. The same procedure as in HTML code. (If you know how 
to program in HTML, you can use it in the .Rmd document)

# xtable

```{r xtab, results = 'asis', message = FALSE, eval = FALSE}
esd <- unlist(lapply(kont, sd, na.rm = TRUE))
emean <- colMeans(kont, na.rm = TRUE)

library(xtable)
print(xtable(summary(kont)), comment = FALSE, include.rownames = FALSE)

```

-->

## stargazer och coxph vs coxreg och ltxCoef3


```{r konfidens, results='asis'}
library(stargazer)
library(survival)
fit <- coxph(Surv(exit, event) ~ sex + age + 
                us.ar + snususe, data = vip)
stargazer(fit, ci = TRUE, apply.coef = exp, ##apply.ci = exp, måste bort!!
          star.char = "", caption = "coxph and stargazer.", report = "vcsp")
```

Not very good output. No p-value for *snususe* and unreliabel ones for the 
reported contrasts. And some noisy output at the end.

```{r coxreg, results = 'asis'}
library(eha)
fit <- coxreg(Surv(exit, event) ~ sex + age + 
                us.ar + snususe, data = vip)
source("R/ltxCoef3.R")
dr <- drop1(fit, test = "Chisq")
ltxCoef3(fit, dr = dr, caption = "coxreg med ltxCoef3.", digits = 3)
```



## Contingency table

Create Table \@ref(tab:twobytwo).

```{r twobytwo}
x <- with(vip, table(sex, snususe))
source("R/tbl.R")
tbl(x, caption = "Sex and snus use.", rownames = TRUE)
```

Table \@ref(tab:twobytwo) is analysed with a $\chi^2$ test:

```{r chisqt}
y <- chisq.test(x)
y
```

As expected, snus habits are very different for men and women. See Table 
\@ref(tab:propts) with relative frequencies (per cent).

```{r propts}
xx <- round(prop.table(x, margin = 1) * 100, 1)
tbl(addmargins(xx, margin = 2),  # Why not 'margin = 1, as above? Check help!
    caption = "Sex and snus use, relative frequencies.",
    rownames = TRUE)
```


## ftable three-way

```{r three}
mytab <- xtabs(~ snususe + sex + alder.VIP, data = vip)
ftable(mytab)
```

Need to investigate further \ldots

# Graphs

## Barplots

for instance, Table  \@ref(tab:propts) is illustrated in Figure \@ref(fig:proptsfig).

```{r proptsfig, fig.cap = "Snus use, men and women."}
barplot(xx, beside = TRUE, col = c("blue", "red"), ylab = "Per cent",
        legend.text = c("Men", "Women"))
abline(h = 0)
```


## lattice

I am using the package *lattice* for this, but it is probably almost as easy to 
do it "by hand", to get control over the layout and "things" \ldots

See Figure \@ref(fig:lattbmi) and so on.


```{r lattbmi, fig.cap = "BMI by sex and snus use.", fig.height = 6}
library(lattice)
library(latticeExtra)
densityplot(~ bmi | sex + snususe, data = vip, plot.points = FALSE)
```

```{r lattinc, fig.cap = "Income by sex and snus use.", fig.height = 6}
library(lattice)
library(latticeExtra)
densityplot(~ income.90 | sex + snususe, data = vip, plot.points = FALSE)
```

Figure \@ref(fig:lattinc) is severely distorted by the fact that a few non-snuffing 
men have fantasy incomes. We restrict income to below 1 million SEK in Figure 
\@ref(fig:lattinc2).

```{r lattinc2, fig.cap = "Income (truncated at 10000) by sex and snus use.", fig.height = 6}
library(lattice)
library(latticeExtra)
densityplot(~ income.90 | sex + snususe, data = vip, plot.points = FALSE,
            xlim = c(0, 10000), xlab = "Truncated income by sex and snus use.")
```



```{r lattage, fig.cap = "Age at start by sex and snus use.", fig.height = 6}
library(lattice)
library(latticeExtra)
densityplot(~ age | sex + snususe, data = vip, plot.points = FALSE)
```

Figure \@ref(fig:lattage) is also distorted by severe heaping in age at start. 
Not so useful.

```{r lattsbt, fig.cap = "Systolic blood pressure (sbt) by sex and snus use.", fig.height = 6}
library(lattice)
library(latticeExtra)
densityplot(~ sbt | sex + snususe, data = vip, plot.points = FALSE)
```

## From scratch

Replacing *lattice*, we can try building from scratch.

```{r fromscratch}
dplo <- function(value, head = "Men, never"){
   ## value = what is plotted
   value <- value[!is.na(value)]
   y <- density(value)
   plot(y$x, y$y, main = head, type = "l", lwd = 1.5, col = "blue",
        ylab = "Density")
   abline(h = 0)
}

```

